### 2014-08-28 standing-variation-paper.tex
latemlified
added more to discussion, and corrected typo in multiple variant section

Over the past ten thousand years malaria resistance alleles have arisen at a number
of genes and spread through human populations in areas where malaria is endemic \citep{Kwiatkowski:05}. 
A number of these alleles appear to represent convergence, with
multiple changes in the same gene. 
For example, there are multiple
changes at the $\beta$-globin gene that confers malaria resistance,
and the sickle cell allele may plausibly have arisen by up to five distinct
changes at the same base pair within Africa
\citep{Flint:98,ralph2010parallel}.
Another particularly impressive case of convergent evolution in response to
selection pressures imposed by malaria are the numerous changes throughout the
X-linked G6PD gene, with upward of 50 polymorphic variants ($>1\%$ frequency) having so far been
described that lower the activity of the enzyme \citep{Howes-g6pd-variants,Minucci-g6pd}. These alleles are now
found at a combined frequency of $\sim$8 \% frequency in malaria endemic countries, rarely
exceeding$\sim$20 \% \citep{Howes-g6pd-preval}. Three malaria
resistance alleles at G6PD are particularly common and relatively well
studied: the $A-$ found in
much of sub-Saharan Africa; the Med allele found in the Mediterranean and middle
east; and the Mahidol allele found Myanmar and Thailand.
The $A-$ and Mahidol alleles have been shown
to be protective against {\it Plasmodium falciparum} and {\it P. vivax} in both
hemizygote males and both heterozygote and homozygotes females \citep{Ruwende-g6pd ,
Louicharoen-g6pd}. Haplotype-based analysis of genetic diversity surrounding
the $A-$, Med, and Mahidol suggest that they have spread in past few
thousand years \citet{tishkoff-g6pd, Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
consistent with the age of other malaria resistance alleles. 
These population genetic analyses suggest that three variants each have a hemizygote/heterozygote 
selection coefficient of $0.05-0.3$ \citep{tishkoff-g6pd
  Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, consistent with
the results of \citet{Ruwende-g6pd} on the basis of the present day increased resistance to malaria of the
$A-$ allele. 
Given such a strong pressure on these alleles they should have risen
quickly to fixation, so their presence at intermediate frequency,
over a broad geographic area, makes it a good candidate for a recently
balanced polymorphism. 
Hemizygous males and homozygous females suffer from G6PD deficiency,
leading to hemolytic anemia when exposed to a variety of compounds, notably
those present in fava beans.

@@

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6pd_Howes_et_al_1475-2875-12-418-4}   %{charlen-by-sd-limit}
\caption{ %
{\bf Map of G6pd-deficiency allele frequencies across Asia.} The pie chart shows
  the frequency of G6pd-deficiency alleles. The size of the pie chart
  indicates the number of G6pd-deficient individuals sampled.
. Countries with
  endemic malaria are colored yellow. Figure taken from
  \citet{Howes-g6pd-variants} \url{http://www.malariajournal.com/content/12/1/418}.

}
\end{center}
\end{figure}

@@

\begin{equation}
    z_0 = \int_0^\infty {\lambda_{0,j} \pi v^2 t^2 - \lambda_j \pi v^2 t^3 / 3} \; \exp \left( - \lambda_{0,j} \pi v^2 t^2 - \lambda_j \pi v^2 t^3 / 3 \right) dt .
\end{equation}

@@

\paragraph{}
Our results touch on an old debate on the evolutionary coherence of
species. 
Mayr and others have argued that species are coherent \citep{} 
as they are united by shared gene flow. However, this arguement has
questioned by a number of authors based on relatively high levels of 
differentiation in many species \citep{}.
\citet{Rieseberg2001,MorjanRieseberg:04} argued that while gene flow
may not be high enough to prevent neutral differentiation, species
should be viewed as being united as levels of gene flow were still
high enough for selected alleles to spread across the entire species.
Our work on geographic convergent adaptation
\citep[see also][]{ralph2010parallel,RalphCoop:14} brings these conclusions partially into
question. Our results suggest that wide-spread species should often
repeatedly adapt in parallel rather than waiting for a single allele
to migrate across the range.

In support of this idea recent selective sweeps are often
geographically restricted \citep{}, but see XXX. This is likely in part due to the
relatively low occurance of species-wide selection pressures \citep{},
but as noted above even when we know of wide-spread selection
pressures (e.g. Malaria) the response is usually convergent across large enough
spatial scales. This view is potentially contradicted by the
introgression of adaptive alleles across species and sub-species
boundaries, suggesting that selected alleles do sometimes spread
despite low migration rates \citep{}. However, at least some of
these cases reflect the introgression of haplotypes of multiple tightly linked beneficial
alleles.

@@

If gene flow rarely rapidly spread selected alleles across
species range, how then do selected alleles ever become fixed within species? 
Repeated bouts of adaptation in particular genomic regions may act to slowly
push a subset of previously selected alleles to fixation across the
species range, through the spread of the genetic backgrounds on which they arise.  
However, it seems likely that this is a slow process compared to the
initial rapid spread of selected alleles.

One potential hypothesis is that many selected alleles only achieve
fixation, not through their own species-wide spread, but rather
through subsequent large-scale
changes in geographic range size induced by extirpation of the species
over part of its range. Such drops in range size may fix alleles, or radically change the
species-wide frequency, of alleles previously restricted to small portion of a species range. 
Range sizes are in constant in flux, with

Similarly the range fragmentation accompanying speciation may act to fix variants within newly formed
species that were polymorphic across ancestral species ranges.

Note that this does not mean that:
Any change in neutral fixation rate.
Any change in

Many DMIs are locally restricted \citep{Cutter:12}.

species range evolution
splitting of species.
\citep{Rosenblum:12}

Futuyma has argued that much of the differentiation within species is \citep{Futuyma:10,FUTUYMA:87}

>>>>

Over the past ten thousand years malaria resistance alleles have arisen at a number
of genes and spread through human populations in areas where malaria is endemic \citep{Kwiatkowski:05}. 
A number of these alleles appear to represent convergence, with
multiple changes in the same gene. 
For example, there are multiple
changes at the $\beta$-globin gene that confers malaria resistance,
and the sickle cell allele may plausibly have arisen by up to five distinct
changes at the same base pair within Africa
\citep{Flint:98,ralph2010parallel}.
Another particularly impressive case of convergent evolution in response to
selection pressures imposed by malaria are the numerous changes throughout the
X-linked G6PD gene, with upward of 50 polymorphic variants ($>1\%$ frequency) having so far been
described that lower the activity of the enzyme \citep{Howes-g6pd-variants,Minucci-g6pd}. These alleles are now
found at a combined frequency of $\sim$8 \% frequency in malaria endemic countries, rarely
exceeding$\sim$20 \% \citep{Howes-g6pd-preval}. Three malaria
resistance alleles at G6PD are particularly common and relatively well
studied: the $A-$ found in
much of sub-Saharan Africa; the Med allele found in the Mediterranean and middle
east; and the Mahidol allele found Myanmar and Thailand.
The $A-$ and Mahidol alleles have been shown
to be protective against {\it Plasmodium falciparum} and {\it P. vivax} in both
hemizygote males and both heterozygote and homozygotes females \citep{Ruwende-g6pd,
Louicharoen-g6pd}. Haplotype-based analysis of genetic diversity surrounding
the $A-$, Med, and Mahidol suggest that they have spread in past few
thousand years \citet{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, 
consistent with the age of other malaria resistance alleles. 
These population genetic analyses suggest that three variants each have a hemizygote/heterozygote 
selection coefficient of $0.05-0.3$ \citep{tishkoff-g6pd,Slatkin-age-est,Saunders-g6pd,Louicharoen-g6pd}, consistent with
the results of \citet{Ruwende-g6pd} on the basis of the present day increased resistance to malaria of the
$A-$ allele. 
Given such a strong pressure on these alleles they should have risen
quickly to fixation, so their presence at intermediate frequency,
over a broad geographic area, makes it a good candidate for a recently
balanced polymorphism. 
Hemizygous males and homozygous females suffer from G6PD deficiency,
leading to hemolytic anemia when exposed to a variety of compounds, notably
those present in fava beans.

@@

\begin{figure}[ht]
\begin{center}
  \includegraphics[width=1.0\textwidth]{G6pd_Howes_et_al_1475-2875-12-418-4}   %{charlen-by-sd-limit}
\caption{ 
Map of G6pd-deficiency allele frequencies across Asia. The pie chart shows
  the frequency of G6pd-deficiency alleles. The size of the pie chart
  indicates the number of G6pd-deficient individuals sampled.
. Countries with
  endemic malaria are colored yellow. Figure taken from
  \citet{Howes-g6pd-variants} \url{http://www.malariajournal.com/content/12/1/418}.}
\end{center}
\end{figure}

@@

\begin{equation}
    z_0 = \int_0^\infty {2 \lambda_{0,j} \pi v^2 t - \lambda_j \pi v^2 t^3 } \; \exp \left(\sum_j - \lambda_{0,j} \pi v^2 t^2 - \lambda_j \pi v^2 t^3 / 3 \right) dt .
\end{equation}

@@

%%Mayr: [t]he steady and high genetic input caused by gene flow is the main factor responsible for genetic cohesion among the populations of a species.

\paragraph{Are species held together by wide-spread selective sweeps?}
Our results touch on an old debate on the evolutionary coherence of
species. Mayr and many others have argued that species are coherent
evolutionary units as they are united by shared gene flow \citep[pages
521–522 ][]{Mayr:SpeciesEvol}. However, this argument has
questioned by a number of authors based on relatively high levels of 
differentiation, and low rates of dispersal, in many species \citep{EhrlichRaven:69,Levin:79}.
\citet{Rieseberg2001,MorjanRieseberg:04} argued that while gene flow
may not be high enough to prevent neutral differentiation, species
should be viewed as being united as levels of gene flow were still
high enough for selected alleles to spread across the entire species \citep{Ellstrand2014s}.
Indeed large scale genotyping and sequencing, along with more
sophisticated methods, are highlighting ever more signals of gene flow
between populations and species \citep{}.
However, our work on geographic convergent adaptation
\citep[see also][]{ralph2010parallel,RalphCoop:14} brings the
conclusion that this gene flow will lead to the rapid spread of
selected alleles partially into
question. Our results suggest that species should often adapt to
wide-spread selection pressures through convergent evolution rather
than waiting for a single allele to migrate across the range over
length-scales determined by our characteristic length.

In support of this idea recent selective sweeps are often
geographically restricted \citep{}, but see XXX. This is likely in part due to the
relatively low incidence of wide-spread selection pressures \citep{},
but as noted above even when we know of wide-spread selection
pressures (e.g. Malaria) the response is usually convergent across large enough
spatial scales. This view is potentially contradicted by the
introgression of adaptive alleles across species and sub-species
boundaries, suggesting that selected alleles do sometimes spread
despite low migration rates \citep{hedrick2013adaptive}. However, at least some of
these cases likely reflect the introgression of haplotype complexes
consisting of multiple tightly linked beneficial
alleles.

@@

\paragraph{How then do substitutions occur?}
It is rare for gene flow to rapidly spread selected alleles across
species range, how then do selected alleles ever become fixed within
species? Repeated bouts of adaptation in particular genomic regions may act 
push a subset of previously selected alleles to fixation across the
species range, through the spread of the genetic backgrounds on which they arise.  
However, it seems likely that this is a slow process compared to the
initial rapid spread of selected alleles.

\paragraph{Speciation and extinction as phases of molecular evolution?}
One potential resolution is that many selected alleles achieve
fixation, not through their own species-wide spread, but rather
through subsequent large-scale changes in geographic range size
induced by extirpation of the species over parts of its range. Such drops in range size may fix, or radically change the
species-wide frequency, of alleles previously restricted to small portion of a species range. 
Also many modes of speciation are proposed to occur through a
geographically-limited subset of populations forming the basis of new species, e.g. the splitting off
of part of the range through a vicariance event or dispersal of a
subset of individuals. As such specation will lead to the geographic assorting of
polymorphic ancestral variation, again acting to fix variants within
newly formed species that were previously polymorphic across ancestral species ranges.

Somewhat along similar lines, Futuyma has argued that much of the adaptive differentiation,
e.g. adaptation to local conditions, within species maybe ephemeral and
subject to loss due to local extinction and the mixing following the collapse of population
structure \citep{Futuyma:10,FUTUYMA:87}. In part this represented a
response to punctuated equilibrium \citep{eldredgegould72}, arguing
that the observation of stasis and rapid anagensis associated with
speciation were consistent with micro-evolution . 
Futuyma argued that despite rapid adaptation over short time-scales, we
may observe morphological stasis in the
fossil record as
much of this adaptation is lost \citep[see
also][]{Eldredge:05}. Furthermore, he suggests that speciation
may act as rachet to prevent the loss of differentiation, acting to
maintain adaptive changes among populations, and prevent their loss by
interbreeding. At face value the rate of species formation seems too low to
contribute to this process. However, as \citep{Rosenblum:12}, and many
others, have argued that in fact the rate of speciation may well be very high,
but that the majority of incipient species do not persist long due to
reabsorption or extinction. Repeated bouts of extinction and
speciation will send waves of alleles
to fixation along particular lineages.

This link between speciation and substitution does not imply that
substitutions should be thought of as being clustered at splits in
inferred phylogenies \citep[see ][for a recent exhange on
this]{Venditti:10, Pennell:14,Venditti:14,Pennell:14B}. 
(Note that neutral substitutions will be to be unaffected by this process, as they
will accumulate in a clock like manner along lineages, as dictated by
the mutation rate, regardless of the geographic details of their
polymorphic stage.)
Firstly, it it likely that splits in
phylogenies are likely only a tiny proportion of all incipient
speciation events because extinction rates may be high
\citep{Rosenblum:12}, such that every lineage has likely past through
many ``speciation events'' in addition to the observed ones.
Secondly, under this view, the origination process of alleles and their fixation process
are uncoupled \citep[in the sense of ][ page XXX]{Gillespiebook} by the geographic structure.
Such that spatial polymorphisms may accumulate gradually, in
geographically restricted populations, between large-scale 
biogeographic events. 
%Thus selected alleles that eventually become a
%substitutions may have arisen long before they reach fixation.

Nor does this view imply that adaptive evolution or speciation is driven by the
shifting balance or genetic revolutions \citep{Wright:32,
  Mayr-genetic-revol:1954}, whereby extreme genetic drift allows
populations to cross fitness valleys and substitute novel epistatic
combinations. While geographic lineage sorting via speciation and
extinction can be thought of as very large-scale genetic drift events,
the initial spread of our alleles is due to selection they are only later sorted

This view suggests that the alleles that substitute along a lineage
may not be those but rather are in part those lucky enough to have
spread across areas not

contributing to

In reality the evolution of range size is not independent of
adaptation. 
Many DMIs are locally restricted \citep{Cutter:12}.

species range evolution
splitting of species.
